---
title: "Clustering Traffic Dataset"
author: "Claire Dudley"
date: "2023-12-03"
output: html_document
---
# Libraries
```{r}
library(tidyverse)
library(cluster)
library(factoextra)
library(dendextend)
library(lubridate)
library(dbscan)
library(clusterCrit)
library(proxy)
library(dtwclust)
library(TSclust)
library(ggdendro)
library(grid)
library(gridExtra)

```

# Dataframe
```{r}
final_df <- read_csv('FINAL_BUILD.csv')
```

# Filtering out NAs
```{r}
final_df <-
  final_df %>%
  filter(!is.na("12:00 AM"))

final_df <- final_df[complete.cases(final_df[ , 4:ncol(final_df)]), ]

final_df$Date <- as.Date(final_df$Date, "%m/%d/%Y")

```

# Removing repeat rows
```{r}
final_df <-
  final_df %>%
  unique()

colnames(final_df)[1] = "location"

final_df <- final_df[, -2]
```


# Selecting feature columns
```{r}
feature_columns <- final_df[, 5:ncol(final_df), drop = FALSE]
```

```{r}
# Example: Impute missing values with mean
feature_columns <- feature_columns %>% mutate_all(funs(ifelse(is.na(.), mean(., na.rm = TRUE), .)))
```

# Scaling
```{r}
z_score <- scale(feature_columns)
threshold <- 3
outliers <- abs(z_score) > threshold
cleaned <- final_df[!apply(outliers, 1, any), ]

feature_columns <- cleaned[, 5:ncol(cleaned), drop = FALSE]
scaled_feature_columns <- as.data.frame(scale(feature_columns))
```


# Distance matrix
```{r}
distance_matrix <- dist(scaled_feature_columns, method = 'euclidean')
```

# Hierarchical Clustering
```{r}
hierarchical_result <- hclust(distance_matrix, method = "ward.D2")
fviz_nbclust(scaled_feature_columns, FUN = hcut, method = "silhouette")
```

```{r}
# Example: Plot hierarchical clustering dendrogram
plot(hierarchical_result, hang = -1, cex = 0.6, main = "Hierarchical Clustering Dendrogram")
rect.hclust(hierarchical_result , k = 2, border = 2:6)
abline(h = 3, col = 'red')

```

cutting dendrogram

```{r}
num_clusters <- 2
cluster_assignment <- cutree(hierarchical_result, k = num_clusters)

table(cluster_assignment, cleaned$location)
```


analyzing cluster membership by location
```{r}
cluster_data <- cbind(cleaned, Cluster = factor(cluster_assignment))

cluster_summary <- cluster_data %>%
  group_by(Cluster, location) %>%
  summarize(across(where(is.numeric), sum)) %>%
  ungroup()

cluster_summary <- cluster_data %>%
  group_by(Cluster, location) %>%
  summarise(count = n()) %>%
  ungroup()

```

analyzing cluster membership by date
```{r}
# Month column
cluster_data$month <- month(mdy(cluster_data$Date))

cluster_summary2 <- cluster_data %>%
  group_by(Cluster, month) %>%
  summarize(across(where(is.numeric), sum)) %>%
  ungroup()

cluster_summary2 <- cluster_data %>%
  group_by(Cluster, month) %>%
  summarise(count = n()) %>%
  ungroup()
```

# Visualizing

Bar plot by location
```{r}
# Assuming cluster_data is your data frame with Cluster and location columns
top_n_locations <- cluster_data %>%
  group_by(Cluster, location) %>%
  summarise(count = n()) %>%
  arrange(Cluster, desc(count)) %>%
  group_by(Cluster) %>%
  top_n(5)  # Adjust the number of top locations to show

# Create bar plot
ggplot(top_n_locations, aes(x = reorder(location, count), y = count, fill = Cluster)) +
  geom_bar(stat = "identity") +
  labs(title = "Top Locations per Cluster", x = "Location", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Bar plot by date
```{r}
# Assuming cluster_data is your data frame with Cluster and date columns
top_n_dates <- cluster_data %>%
  group_by(Cluster, month) %>%
  summarise(count = n()) %>%
  arrange(Cluster, desc(count)) %>%
  group_by(Cluster) %>%
  top_n(5)  # Adjust the number of top dates to show

# Create bar plot
ggplot(top_n_dates, aes(x = reorder(month, count), y = count, fill = Cluster)) +
  geom_bar(stat = "identity") +
  labs(title = "Top Months per Cluster", x = "Month", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Evaluation
ran through each linkage method, results were written in below
```{r}
silhouette_score <- silhouette(cluster_assignment, distance_matrix)
mean(silhouette_score[, 'sil_width'])
```
ward: 0.441
average: 0.456
complete:0.441
single: 0.226




# DBScan

```{r}
eps_values <- seq(0.1, 1.0, by = 0.1)
MinPts_values <- c(3, 5, 10, 15)

# Running a grid search to fine tune parameters
results <- data.frame()

for (eps in eps_values) {
  for (MinPts in MinPts_values) {
    dbscan_result <- dbscan(scaled_feature_columns, eps = eps, MinPts = MinPts)
    
    # if cluster formed
    if (length(unique(dbscan_result$cluster)) > 1) {
      
      dist_matrix <- proxy::dist(as.matrix(scaled_feature_columns))
      
      silhouette_score <- silhouette(dbscan_result$cluster, dist_matrix)
      
      # Value needed for silhouette score
      sil_width <- silhouette_score[, "sil_width"]
      
      avg_silhouette <- mean(sil_width)
      
      results <- rbind(results, data.frame(eps = eps, MinPts = MinPts, silhouette = avg_silhouette))
    } else {
      # HAndle
      warning(paste("No clusters found for eps =", eps, "and MinPts =", MinPts))
    }
  }
}
```


```{r}
eps <- 0.3
minPts <- 3

dbscan_res <- dbscan(scaled_feature_columns, eps = eps, minPts = minPts)

```



# Evaluation
```{r}
dist_matrix <- proxy::dist(as.matrix(scaled_feature_columns))
silhouette_score <- silhouette(dbscan_res$cluster, dist_matrix)
mean(silhouette_score[, "sil_width"])
```
0.0913


# Comparing DBScan and Hierarchical Clustering
```{r}
# Add the cluster labels to the original data frame
cleaned$dbcluster <- dbscan_res$cluster
cleaned$hierarchical_cluster <- factor(cluster_assignment)
```

```{r}
cluster_data <- cbind(cleaned, dbscan_clust = dbscan_res$cluster)
cluster_data <-
  cluster_data %>%
  mutate(Cluster = as.factor(dbscan_clust + 1))

cluster_summary <- cluster_data %>%
  group_by(Cluster, location) %>%
  summarize(across(where(is.numeric), sum)) %>%
  ungroup()

cluster_summary <- cluster_data %>%
  group_by(Cluster, location) %>%
  summarise(count = n()) %>%
  ungroup()

```

analyzing cluster membership by date
```{r}
# Month column
cluster_data$month <- month(ymd(cluster_data$Date))

cluster_summary2 <- cluster_data %>%
  group_by(Cluster, month) %>%
  summarize(across(where(is.numeric), sum)) %>%
  ungroup()

cluster_summary2 <- cluster_data %>%
  group_by(Cluster, month) %>%
  summarise(count = n()) %>%
  ungroup()
```

Bar plot by location
```{r}
# Assuming cluster_data is your data frame with Cluster and location columns
top_n_locations <- cluster_data %>%
  group_by(Cluster, location) %>%
  summarise(count = n()) %>%
  arrange(Cluster, desc(count)) %>%
  group_by(Cluster) %>%
  top_n(5)  # Adjust the number of top locations to show

# Create bar plot
ggplot(top_n_locations, aes(x = reorder(location, count), y = count, fill = Cluster)) +
  geom_bar(stat = "identity") +
  labs(title = "Top Locations per Cluster", x = "Location", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Bar plot by date
```{r}
# Assuming cluster_data is your data frame with Cluster and date columns
top_n_dates <- cluster_data %>%
  group_by(Cluster, month) %>%
  summarise(count = n()) %>%
  arrange(Cluster, desc(count)) %>%
  group_by(Cluster) %>%
  top_n(5)  # Adjust the number of top dates to show

# Create bar plot
ggplot(top_n_dates, aes(x = reorder(month, count), y = count, fill = Cluster)) +
  geom_bar(stat = "identity") +
  labs(title = "Top Months per Cluster", x = "Month", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Evaluation

# KMeans

Elbow method
```{r}
fviz_nbclust(scaled_feature_columns, FUN = kmeans, method = "silhouette")
```
so 2 clusters

```{r}
kmeans_result <- kmeans(feature_columns, centers = 2)
```



```{r}
fviz_cluster(kmeans_result, data = scaled_feature_columns,
             palette = c("#2E9FDF", "#00AFBB", "#E7B800"), 
             geom = "point",
             ellipse.type = "convex", 
             ggtheme = theme_bw()
             )
```

```{r}
silhouette_score <- silhouette(kmeans_result$cluster, distance_matrix)
mean(silhouette_score[, 'sil_width'])
```
0.441



